Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages

### 2018-02-01 manuscript/manuscript.tex
added figure for phylobayes

\begin{abstract}
\noindent \skhcomment{$\leq$ 250 words, currently 164}   
Molecular phylogenetics is often used to estimate the time since the divergence of modern gene sequences.
Such phylogenetic techniques often estimate substantially shallower divergence times than other methods.
For instance, in the case of viruses there is independent evidence that molecular phylogenetics can underestimate deep divergence times.  
This discrepancy is thought to be caused in part by inadequate models of purifying selection leading to branch-length underestimation.
Here we show that substitution models informed by experimental measurements of site-specific purifying selection lengthen the branches on phylogenies of influenza virus hemagglutinin.
This deepening of branch lengths is due to better modeling how site-specific amino-acid preferences affect the stationary state of the substitution models, and is independent of the branch-length-extension due to modeling site-to-site variation in substitution rate.
We show that the deepening of branch lengths from experimentally informed site-specific substitution models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
However, the improvements from these site-specific models are limited by the inherent tension between the enhanced accuracy of accounting for site-specific amino-acid preferences and the fact that these preferences shift over long evolutionary times.
Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state as well as the substitution rate when estimating deep divergence times. 
\end{abstract}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/model_feature.pdf}}
\caption{\label{fig:model_feature}
\textbf{Comparison of substitution model features.}
Site-specific amino-acid profiles and $\Gamma$-distributed rate variation are both substitution model features which have been shown or theorized to lengthen branches. 
The models YNGKP M0, YNGKP M5, ExpCM, and ExpCM+$\Gamma\omega$ represent all possible combinations of these two features. 
Blue indicates presence and white indicates absence of a feature. 
from \ref{tab:wsn_low_params}
}
\end{figure}

\begin{itemize}
\item There are different strategies to account for purifying selection and site-to-site rate variation. 
\begin{itemize}
\item $\Gamma\omega$ rate variation 
\begin{itemize}
\item single parameter controlling relative rate of non-synonymous to synonymous substitutions can vary across sites following some statistical distribution (usually $\Gamma$)
\item very common 
\item practical implementation - discretized distribution with the mean of $K$ bins. 
\end{itemize}
\item stationary state
\begin{itemize}
\item Explicitly model each amino-acid frequency at each site in the protein 
\item a more mechanistic description, no arbitrary statistical distribution 
\item practical limitation - how do you estimate these parameters without overfitting the data? We use experimental measurements to define the parameters \textit{a priori}
\end{itemize}
\end{itemize}
\item We can compare the effect of a model feature ($\Gamma\omega$ rate variation or non-uniform stationary state) while controlling the presence/absence of the other feature using models from the ExpCM and GY94 families. 
\end{itemize}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/decay.pdf}}
\caption{\label{fig:decay}
\textbf{Effect of stationary state and rate variation on long branch estimation.}
The expected pairwise identity trajectories were calculated using \ref{eq:f} and models described in \ref{tab:wsn_low_params}.
The trajectories of the YNKGP M0 (blue) and YNGKP M5 (green) do not vary from panel to panel because neither model is site-specific. 
The deviation in trajectory of the ExpCM (red) from the YNGKP M0 (blue) increases from left to right as the mutational constraint of the amino-acid profile increases (logoplots, above). 
The deviation in trajectory of the YNGKP model with a site-specific $\omega$ value inferred from the ExpCM (yellow, \ref{eq:w_r}) is also positively correlated with the constraint of the site-specific amino-acid profiles but the effect size is smaller. 
}
\end{figure}

\begin{itemize}
\item Substitution models are only useful for inferring branch lengths when they can accurately transform sequence divergence to time (branch lengths)
\item Every substitution model has a saturation point where this temporal signal is lost.  
\begin{itemize}
\item In the long tails the expected sequence divergence remains constant as time increases. 
\item The saturation point will affect long branches more than short branches.
\end{itemize}
\item Different model features (\ref{fig:model_feature}) can affect \textit{how long} it takes to reach the saturation point and the \textit{sequence divergence} at the saturation point 
\begin{itemize}
\item Adding $\Gamma\omega$ rate variation affects the ``path" to the saturation point but affects the sequence divergence at the saturation point minimally. Importantly, the saturation point is identically for every site in the protein. (green and blue)
\item Adding a non-uniform stationary state creates a unique saturation point for each site in the protein (red). Sites which can tolerate many mutations have saturation points which are quite similar to the GY94 saturation point. Constrained sites show a dramatic difference. At a constrained site, ExpCM will estimate a longer branch than the GY94 model because the overall sequence divergence at the site is expected to be small. 
\item This difference cannot be recapitulated by more complex modeling of the $\omega$ parameter. We can transform the GY94 model into a  site-specific model by inferring a unique $\omega$ value from the ExpCM for each site in the protein. This model (yellow) takes longer to reach the saturation point than the GY94 model but ultimately has the same divergence level at the saturation point. 
\end{itemize}
\item No matter how ``well" you model site-to-site rate variation, you will underestimate \textit{very} long branches with a uniform stationary state. 
\end{itemize}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.9\textwidth]{figures/empirical_trees}}
\caption{\label{fig:empirical_trees}
\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
\jdbcomment{Keep branch lengths in substitutions per site. Make a panel A show a prefs snippet, B H1 trees, C H3 trees, D H1+H3. Maybe increase very small font sizes a bit.}
The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H1 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H1 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H1 focal sequence (black triangle).
}
\end{figure}

@@

\subsection*{texttt{phylobayes}}

>>>>

\begin{abstract}
\noindent \skhcomment{$\leq$ 250 words, currently 217}   
Molecular phylogenetics is often used to estimate the time since the divergence of modern gene sequences.
Such phylogenetic techniques often estimate substantially shallower divergence times than other methods.
For instance, in the case of viruses there is independent evidence that molecular phylogenetics can underestimate deep divergence times.  
This discrepancy is thought to be caused in part by inadequate models of purifying selection leading to branch-length underestimation.
Here we show that substitution models informed by experimental measurements of site-specific purifying selection lengthen the branches on phylogenies of influenza virus hemagglutinin.
This deepening of branch lengths is due to better modeling how site-specific amino-acid preferences affect the stationary state of the substitution models, and is independent of the branch-length-extension due to modeling site-to-site variation in substitution rate.
We show that the deepening of branch lengths from experimentally informed site-specific substitution models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
However, the improvements from these site-specific models are limited by the inherent tension between the enhanced accuracy of accounting for site-specific amino-acid preferences and the fact that these preferences shift over long evolutionary times.
Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state as well as the substitution rate when estimating deep divergence times. 
\end{abstract}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/model_feature.pdf}}
\caption{\label{fig:model_feature}
\textbf{Comparison of substitution model features.}
(A) Substitution model stationary states can either be identical at every site in the protein or allow to vary from site to site. 
(B) The relative rate of non-synonymous change, $\omega$, can be defined as one, gene-wide average or allowed to vary following some statistical distribution. 
To ensure computational tractability, $\omega$ takes on the mean value of $K$ bins ($K=4$ shown). 
(C) Substitution models can both, one, or neither of these features. 
We used models from the GY94 and the ExpCM families. 
}
\end{figure}

It is well known that protein sites evolve under different pressures. 
Some sites appear to tolerate many amino acids while other sites appear to tolerate only a select few. 
Therefore, it is important to use substitution models which are able to describe this purifying selection, and the resulting site-to-site rate variation, when modeling protein evolution.

The standard strategy to account for purifying selection is to allow the rate of non-synonymous change to vary among sites following some statistical distribution. 
This variation is commonly implemented by allowing the rate parameter, $\omega$, to follow a $\Gamma$-distribution ($\Gamma\omega$).
In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ is assigned to the mean of each bin and the site likelihood is the average likelihood given each of the $\omega$ values. 
Under this model, some sites would have a higher rate of non-synonymous change than others but this rate would be agnostic to the amino-acid identities themselves. 
All non-synonymous changes would be treated as equally.

Another strategy to account for purifying selection is to define a substitution model with a site-specific stationary state. 
Such models describe the evolution at each site uniquely and independently of the other sites. 
For example, models from the Halpern-Bruno mutation-selection family explicitly describe the frequency of each amino acid at each site. 
However, these more mechanistic models, without an arbitrary statistical distribution, come at a cost in the form of an increased number of parameters. 
When using a mutation-selection model is crucial to invoke a strategy to avoiding overfitting the data. 
We chose to define the amino-acid frequency parameters \textit{a priori} using measurements from a high-throughput functional assay called deep mutational scanning, Experimentally Informed Codon Models (ExpCMs).

These two strategies, $\Gamma\omega$ rate variation and non-uniform stationary states, are not truly distinct. 
Differences in stationary state manifest as differences in non-synonymous rates across sites \citep{spielman2015relationship}. 
Moreover, these two strategies are not mutually exclusive. 
Technically, substitution models can have a non-uniform stationary state and $\Gamma\omega$ rate variation and biologically, these models can account for both strong purifying and strong positive selection.

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/decay.pdf}}
\caption{\label{fig:decay}
\textbf{Effect of stationary state and rate variation on long branch estimation.}
The expected pairwise amino-acid identity for five sites in influenza hemagglutinin (HA) for four different substitution models. 
The logoplots show the HA amino-acid preferences from a deep mutational scan performed by \cite{doud2016accurate}. 
The implied dN/dS value was calculated from the ExpCM following \cite{spielman2015relationship}.
}
\end{figure}

Along a single branch, a substitution model transforms sequence divergence to branch length, which under a molecular clock assumption is proportional to time. 
However, every substitution model has a stationary state, the expected sequence divergence given a very long time, at which the relationship between sequence divergence and time is lost. 
The stationary state is represented by the long ``tails" in \ref{fig:decay}, where the expected sequence divergence remains constant as time increases.

A substitution model's stationary state is defined by the parameters in the model and can vary from model to model. 
Some model features affect how long it takes a model to reach its stationary state, whereas other features affect the stationary state itself. 
Adding $\Gamma\omega$ rate variation affects the ``path" to the stationary state but does not affect the stationary state itself. 
In addition, while $\Gamma\omega$ accounts for site-to-site rate variation, the resulting stationary state is identical at each site in the protein. 
Under these models with uniform stationary states, a mutationally constrained site and a site which tolerates many mutations will have the sequence divergence given a very long time. 
Models with non-uniform stationary states, by definition, acknowledge that differences in sites will persist even at a very long period of time. 
The difference in branch length estimation will be most dramatic at constrained sites in the protein. 
At a constrained site, the ExpCM will estimate a longer branch than the GY94 model because the overall sequence divergence at that site is expected to be small.

This difference in branch length estimation cannot be recapitulated by more complex modeling of the $\omega$ parameter. 
We transformed the GY94 model into a site-specific model by inferring a unique $\omega$ value from the ExpCM, following the work of \cite{spielman2015relationship}, for each site in the protein. 
This model takes longer to reach stationary state than the GY94 model ultimately has the same divergence level at stationary state. 
As this model shows, no matter how ``well" a model accounts for site-to-site rate variation, it will underestimate long branches with a uniform stationary state model.

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.9\textwidth]{figures/empirical_trees.pdf}}
\caption{\label{fig:empirical_trees}
\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
\jdbcomment{Keep branch lengths in substitutions per site. Make a panel A show a prefs snippet, B H1 trees, C H3 trees, D H1+H3. Maybe increase very small font sizes a bit.}
The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H1 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H1 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H1 focal sequence (black triangle).
}
\end{figure}

\begin{table}[t!]
\caption{\label{tab:empirical_data}
Model comparison for Fig.~\ref{fig:empirical_trees}.}
     \begin{tabular}{ccccc}
        \hline
          Model & $\Delta$AIC & Log Likelihood & Parameters\\ \hline
       	ExpCM + $\Gamma\omega$ (average) & 0.00 & -48751.16 & $\omega$ = (0.19,  0.50,  0.90,  1.86), $\beta=1.70$, $\kappa=3.73$\\
	ExpCM (average) & 949.70 & -49227.01 & $\beta=1.78$, $\kappa=3.40$, $\omega=0.15$\\
	ExpCM + $\Gamma\omega$ (H1) & 1306.44 & -49404.38  & $\omega$ = (0.13 ,  0.44,  0.91,  2.16), $\beta=1.12$, $\kappa=3.64$\\
	ExpCM + $\Gamma\omega$ (H3) & 1737.40 & -49619.86 & $\omega$ = (0.09,  0.33,  0.72,  1.77), $\beta=1.28$, $\kappa=3.71$\\
	ExpCM (H1) & 2555.58 & -50029.95 & $\beta=1.22$, $\kappa=3.24$, $\omega=0.13$\\
	ExpCM (H3) & 3196.50 & -50350.41 & $\beta=1.45$, $\kappa=3.38$, $\omega=0.12$\\
	GY94 + $\Gamma\omega$ & 4719.34 & -51105.83 & $\omega$ = (0.00,  0.03,  0.08,  0.26), $\kappa=3.18$\\
	GY94 & 7624.64 & -52559.48  & $\kappa=2.89$, $\omega=0.07$\\
      \end{tabular}
\end{table}

@@

\subsection*{\texttt{phylobayes}}

\begin{figure}[H]
\centerline{\includegraphics[width=0.5\textwidth]{figures/phylobayes.pdf}}
\caption{\label{fig:phylobayes}
\textbf{Comparison of ExpCM and phylobayes}
}
\end{figure}
